Informative Autonomous Sampling of Oceanographic Variables Using Joint Excursion Sets

### 2019-10-14 main_new.tex
Minor edits in Intro+near Fig 10. Trygve; fix p(x;y) in legends of Fig 10c,d

\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{t1,t2}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{t3}, \ead[label=e2]{jo.eidsvik@math.ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{t4}, \ead[label=e3]{david.ginsbourger@idiap.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{t2,t5,t6}. \ead[label=e4]{kanna.rajan@ntnu.no}}

\affiliation[t1]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway.} 
\affiliation[t2]{Centre for Autonomous Marine Operations and Systems (AMOS).}
\affiliation[t3]{Department of Mathematical Sciences, NTNU.}
\affiliation[t4]{Uncertainty Quantification and Optimal Design group, Idiap Research Institute, Martigny, Switzerland.}
\affiliation[t5]{Department of Engineering Cybernetics, NTNU}
\affiliation[t6]{Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

@@

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand dynamic processes in the water-column, the
  combination of statistics and autonomous robotics provide new
  opportunities for studying experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal processes by providing informative descriptions of spatial
  characteristics of ocean phenomena. Specifically, we define a design
  criterion based on reducing uncertainty in the joint excursions of
  random fields, and derive analytical expressions for the expected
  Bernoulli variance reduction in the excursion set of spatial
  temperature and salinity variables, providing a description of
the boundary uncertainty that can be used to prioritize sampling efforts at locations that are
ambiguous, making exploration more effective. We show how these can be
  applicable in static and sequential designs, building on myopic and
  non-myopic sampling strategies. We use simulations to study the
  properties and differences between state-of-the-art approaches,
  followed by results from field deployments with an autonomous
  underwater vehicle as part of a case study mapping the boundary of a
  river plume. The results demonstrate the potential of combining statistical methods and robotic platforms, and the abilities of the derived methods to effectively inform and execute data-driven environmental sampling.

@@

\item Full numerical ocean models cannot provide accurate results
  online if run on robotic sensing platforms, as onboard computers
  cannot deliver the computational power required, hence statistical
  proxy models of the environment must be used for doing inference
  about where to sample.

\item As a priori information about the state of the ocean is
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken in-situ. Recently
  acquired information can be used to resolve this by assimilating up-to-date data into statistical models that
  can be used to inform decisions on where to sample
  sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.

\item For sampling problems related to environmental sensing, the
  number of choices (i.e. locations, trajectories, and candidate
  designs) makes the problem combinatorially large, creating a
  trade-off between optimization (finding the most resource-efficient
  design to collect necessary data), and computability (arriving at a
  solution in reasonable time). To successfully resolve features, this
  trade-off has to be considered in development and practice.

@@

In this work, we build on joint excursion sets (ES) and excursion
probabilities (EP) for the spatial design of experiments. The goal is
to design sampling strategies for reducing the spatially-integrated
Bernoulli variance (IBV) in the ES indicators. Our main contributions
include: i) extending the IBV criteria to multivariate
GPs, thereby allowing the evaluation to include multiple environmental
parameters such as temperature and salinity; ii) developing methods
for information-based environmental sensing and autonomous data
collection; and iii) demonstrating the approach in the real-world
scenario of sampling a river plume using an AUV.

@@

Our objective is to leverage statistical tools to characterize a river
plume, focusing on spatial separation of cold freshwater from the
river and warmer saline waters of the fjord. ES and EPs are useful
starting points to measure the ability to classify water masses. The
goal of experimental sampling is to improve this
characterization, as quantified via changes in the EPs.

@@

For the spatial dependence among variables, we assume a separable
correlation function with the same decay for salinity and temperature,
which is consistent with expectations for a river plume where the
decay is related to common water masses (Section
\ref{sec:case_study}). In the resulting function, we have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and
$\delta_{ij}=0$ otherwise. Moreover, we assume stationary isotropic
random fields a priori where the correlation function $\rho(\bx,\bx')$
solely depends on $\bx$ and $\bx'$ via Euclidean distance
$h=\sqrt{||\bx-\bx'||^2}$. With data and prior knowledge, one could
possibly fit and estimate parameters of non-stationary or
non-isotropic covariance functions and more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

@@

The solution to the second part of Eq. \eqref{eq:22} can be found by
evaluating a four-variable cumulative distribution function associated
with Eq. (\ref{part1:phi4}), accounting for the mean, variance and
covariances of $\bZ_1$, $\bZ_2$ and $\bm$ in the appropriate linear
combinations. Using vector-matrix formulations, we define the linear
combinations:
\begin{equation}
\label{Vcomb}
    \bV=\left[
    \begin{array}{cccccc}
        a_T & 0 & 0 & 0 & 1 & 0 \\
         0 & a_S & 0 & 0 & 0 & 1 \\
         0 & 0 & a_T & 0 & 1 & 0\\
         0 & 0 & 0 & a_S & 0 & 1\\
    \end{array}
    \right] 
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right),
\end{equation}
which in turn then becomes:
{\small
\begin{eqnarray}
\label{part1:phi4_2}
&&P \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}}, Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right)\\
&=&  \Phi_4 
\begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S \\
t_T \\
t_S
\end{bmatrix};
E( \bV),Var(\bV)
\end{pmatrix} \nonumber.
\end{eqnarray}
}
The multivariate cumulative probabilities in two and four dimensions
are computationally efficient using methods developed by
\cite{genz2009computation}.  In the end, the first part in
Eq. (\ref{two_parts0}) and the second in Eq. (\ref{part1:phi4}) are
computed for each location on the grid and summed over
$\bx \in \mathcal{M}_g$, to get the design evaluation criterion in
Eq. (\ref{sur}).

@@

%\subfigure[ES for Survey 1]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_1.pdf}\label{fig:res3}}\hspace{0.4cm}
%\subfigure[ES for Survey 2]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_4.pdf}\label{fig:res4}}
\caption{Results from mapping the Nidelva river. \ref{fig:map} Show an
  overview of the survey area overlaid with the AUV path in black and
  dashed line. Note the shaded region indicating the usual frontal
  region. \ref{fig:res_both} Presents the collected temperature data
  as colored trails. Note waypoint 5 (WP5) which indicate where the
  two surveys deviate. \ref{fig:res1} and \ref{fig:res2} Shows the
  collected salinity data (as a colored trail) overlaid on the final
  EP, which indicate the AUVs statistical impression of the front. For
  both missions the temperature and salinity data correspond with the
  EP front indication.}
\label{fig:results}
\end{figure*}

@@

Two missions, Surveys 1 and 2, were run successively from 11:00 AM to
01:00 PM, with a short break in between. The resulting path of the
selected waypoints are shown in the map in Fig. \ref{fig:map}, both
within the expected frontal region (shaded pink). The recorded
temperatures are shown as colored trails in Fig. \ref{fig:res_both},
clearly indicating the temperature difference between fjord and river
waters. The salinity data are then shown separately, overlaid with the
estimated EP for each survey in Fig. \ref{fig:res1} and
\ref{fig:res2}.

Both surveys successfully estimated and navigated the separation zone,
crossing the frontal boundary multiple times. As conditions changed
slightly between the two deployments, the resulting path (after
waypoint 5) is shown to deviate. Survey 1 continued northwards,
tracking the north-eastern part of the front, while Survey 2 turned
west, mapping the south-western part.

The final predictions of the front location given in
Fig. \ref{fig:res1} and \ref{fig:res2} as a dashed line, correspond
with one another, yielding a picture of the front gradually bending
off toward the north east. The amount of exploration done by Survey 1
is greater than Survey 2. In Survey 1, the AUV obtained more detail by
going north, while Survey 2, coming close to the survey area borders
in the south-western corner, obtained a poorer understanding of the
northern parts. A longer planning horizon would likely identify and
discourage such choices.

>>>>

\begin{aug}
\author{\fnms{Trygve Olav} \snm{Fossum}\thanksref{t1,t2}, \corref{} \ead[label=e1]{trygve.o.fossum@ntnu.no}}
\author{\fnms{Jo} \snm{Eidsvik}\thanksref{t3}, \ead[label=e2]{jo.eidsvik@math.ntnu.no}}
\author{\fnms{David} \snm{Ginsbourger}\thanksref{t4,t5}, \ead[label=e3]{david.ginsbourger@idiap.ch}}
\and
\author{\fnms{Kanna} \snm{Rajan}\thanksref{t2,t6,t7}. \ead[label=e4]{kanna.rajan@ntnu.no}}

\affiliation[t1]{Department of Marine Technology, The Norwegian University of Science and Technology (NTNU), Trondheim, Norway.} 
\affiliation[t2]{Centre for Autonomous Marine Operations and Systems (AMOS).}
\affiliation[t3]{Department of Mathematical Sciences, NTNU.}
\affiliation[t4]{Uncertainty Quantification and Optimal Design group, Idiap Research Institute, Martigny, Switzerland.}
\affiliation[t5]{Department of Mathematics and Statistics, Uncertainty of Bern, Switzerland.}
\affiliation[t6]{Department of Engineering Cybernetics, NTNU}
\affiliation[t7]{Underwater Systems and Technology Laboratory, Faculty of Engineering, University of Porto, Portugal.}

@@

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand processes in the water-column, the
  combination of statistics and autonomous robotics provide new
  opportunities for experimental designs. In this work we develop
  methods for efficient spatial sampling applied to the mapping of coastal processes by providing informative descriptions of spatial
  characteristics of ocean phenomena. Specifically, we define a design
  criterion based on reducing uncertainty in the joint excursions of vector-valued Gaussian random fields, and derive tractable expressions for the expected
  Bernoulli variance reduction in such a framework. We demonstrate how these provide a description of
the boundary uncertainty in temperature and salinity variables, which is used to prioritize sampling efforts at locations that are ambiguous, making exploration more effective by building on myopic and
  non-myopic strategies. We use simulations to study the
  properties and differences between state-of-the-art approaches,
  followed by results from field deployments with an autonomous
  underwater vehicle as part of a case study mapping the boundary of a
  river plume. The results demonstrate the potential of combining statistical methods and robotic platforms to effectively inform and execute data-driven environmental sampling.

@@

\item Full numerical ocean models cannot provide accurate results
  online if run on robotic sensing platforms, as onboard computers
  cannot deliver the computational power required, hence statistical
  proxy models of the environment must be used for learning where to sample.

\item With limited available information about the state of the ocean, there is substantial value in reacting to
  information obtained from measurements taken in-situ. This
  acquired information must be assimilated into statistical models that
  can be used to inform decisions on where to sample
  sequentially.%\emph{in-situ}; this is usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.

\item For sampling problems related to environmental sensing, the
  number of choices (i.e. locations, trajectories, and candidate
  designs) is enormous, creating a
  trade-off between optimization (finding the most resource-efficient
  design to collect necessary data), and computability (arriving at a
  solution in reasonable time). To successfully resolve features, this
  trade-off has to be considered in development and practice.

@@

In this work, we build on joint excursion sets (ES) and excursion
probabilities (EP) for the spatial design of experiments. The goal is
to design sampling strategies for for improving the estimation of joint ES with respect to temperature and salinity by reducing the spatially integrated
Bernoulli variance (IBV) in the ES indicators. Our main contributions
include: i) extending the IBV criteria to multivariate
GPs, thereby allowing the evaluation to include multiple environmental
parameters such as temperature and salinity; ii) developing methods
for information-based environmental sensing and autonomous data
collection; and iii) demonstrating the approach in the real-world
scenario of sampling a river plume using an AUV.

@@

Our objective is to leverage statistical approaches to characterize a river
plume, focusing on spatial separation of cold freshwater from the
river and warmer saline waters of the fjord. ES and EPs are useful
starting points to measure the ability to classify water masses. The
goal of experimental sampling is to improve this
characterization, as quantified via changes in the EPs.

@@

For the spatial dependence among variables, we assume a separable
correlation function with the same decay for salinity and temperature,
which is consistent with expectations for a river plume where the
decay is related to common water masses (Section
\ref{sec:case_study}). In the resulting function, we have:
\begin{equation}\label{gp_corr}
\mbox{Cov}(\xi_i(\bx),\xi_j(\bx')) = \rho(\bx,\bx') \sigma_i \sigma_j (\gamma +(1-\gamma )\delta_{ij}), \\
\end{equation}
where for $i,j \in {T,S}$ and $\delta_{ij}=1$ if $i=j$, and
$\delta_{ij}=0$ otherwise. Moreover, we assume stationary isotropic
random fields where the correlation function $\rho(\bx,\bx')$
solely depends on $\bx$ and $\bx'$ via Euclidean distance
$h=\sqrt{||\bx-\bx'||^2}$. With data and prior knowledge, one could
possibly fit and estimate parameters of non-stationary or
non-isotropic covariance functions and more complex multivariate
spatial covariance functions
\citep{gneiting2010matern,genton2015cross}; however, that is left for
future work.

@@

The solution to the second part of Eq. \eqref{eq:22} can be found by
evaluating a four-variable cumulative distribution function associated
with Eq. (\ref{part1:phi4}), accounting for the mean, variance and
covariances of $\bZ_1$, $\bZ_2$ and $\bm$ in the appropriate linear
combinations. Using vector-matrix formulations, we define the linear
combinations:
\begin{equation}
\label{Vcomb}
    \bV=\left[
    \begin{array}{cccccc}
        a_T & 0 & 0 & 0 & 1 & 0 \\
         0 & a_S & 0 & 0 & 0 & 1 \\
         0 & 0 & a_T & 0 & 1 & 0\\
         0 & 0 & 0 & a_S & 0 & 1\\
    \end{array}
    \right] 
    \left(
    \begin{array}{cccccc}
         Z_{1,T} \\
         Z_{1,S} \\
         Z_{2,T} \\
         Z_{2,S} \\
         m_{T} \\
         m_{S}
    \end{array}
    \right),
\end{equation}
which in turn then becomes:
{\small
\begin{eqnarray}
\label{part1:phi4_2}
&&P \left( Z_{1,T} \leq \frac{t_T-m_{T}}{a_{T}}, 
Z_{1,S} \leq \frac{t_S-m_{S}}{a_{S}}, Z_{2,S} \leq \frac{t_T-m_{T}}{a_{T}}, Z_{2,S} \leq \frac{t_S-m_{S}}{a_{S}} \right)\\
&=&  \Phi_4 
\begin{pmatrix} 
\begin{bmatrix} t_T\\
t_S \\
t_T \\
t_S
\end{bmatrix};
E( \bV),Var(\bV)
\end{pmatrix} \nonumber.
\end{eqnarray}
}
The multivariate cumulative probabilities in two and four dimensions 
are available using methods developed by
\cite{genz2009computation}.  In the end, the first part in
Eq. (\ref{two_parts0}) and the second in Eq. (\ref{part1:phi4}) are
computed for each location on the grid and summed over
$\bx \in \mathcal{M}_g$, to get the design evaluation criterion in
Eq. (\ref{sur}).

@@

%\subfigure[ES for Survey 1]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_1.pdf}\label{fig:res3}}\hspace{0.4cm}
%\subfigure[ES for Survey 2]{\includegraphics[height=0.41\textwidth]{Figures/field-trials/ep_4.pdf}\label{fig:res4}}
\caption{Results from mapping the Nidelva river. \ref{fig:map} shows an
  overview of the survey area overlaid with the AUV paths in black and
  dashed line. The shaded region indicates the usual frontal
  region. \ref{fig:res_both} presents the collected temperature data
  as colored trails. Waypoint 5 (WP5) is where the
  two surveys deviate. \ref{fig:res1} and \ref{fig:res2} show the
  collected salinity data in the two surveys (as colored trails) overlaid on the final
  EPs, which indicate the AUVs statistical impression of the front in the surveys. For
  both missions the temperature and salinity data correspond with the
  EP front indication.}
\label{fig:results}
\end{figure*}

@@

Two missions, Surveys 1 and 2, were run successively from 11:00 AM to
01:00 PM, with a break in between. The resulting path of the
selected waypoints are shown in the map in Fig. \ref{fig:map}, both
within the expected frontal region (shaded pink). The recorded
temperatures in the two surveys are shown as colored trails in Fig. \ref{fig:res_both},
clearly indicating the temperature difference between fjord and river
waters. The salinity data are then shown separately, overlaid with the
estimated EPs for each survey in Fig. \ref{fig:res1} and
\ref{fig:res2}.

Both surveys successfully estimated and navigated the separation zone,
crossing the frontal boundary multiple times. As conditions changed
slightly between the two deployments, the resulting paths (after
waypoint 5) are shown to deviate. Survey 1 continued northwards,
tracking the north-eastern part of the front, while Survey 2 turned
west, mapping the south-western part.

The final predictions of the front location, represented by conditional EPs in
Fig. \ref{fig:res1} and \ref{fig:res2} as dashed lines, correspond
with one another. In both surveys they yield a picture of the front being a bit to the west in the southern parts of the domain and gradually bending off toward the north east. The amount of exploration done by Survey 1
is greater than Survey 2. In Survey 1, the AUV obtained more detail by
going north from waypoint 5, while Survey 2, coming close to the survey area borders
in the south-western corner, obtained a poorer understanding of the
northern parts. A look-ahead strategy might identify and
discourage such choices.
